---
title: "Homework II - Basketball"
author: "Arnau Garcia"
date: "2024-03-26"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Homework II - Basketball

## Section a):

We collect the average of points per game in the last twenty years of the NBA and the ACB, and we multiply by 2 because we are interested in the total amount of points scored per game.

```{r}
nba<- c(114.5, 114.7, 110.6, 112.1, 111.8, 111.2, 106.3, 105.6, 102.7, 
        100.0, 101.0, 98.1, 96.3, 99.6, 100.4, 100, 99.9, 98.7, 97.0, 97.2, 93.4)  
acb <- c(83.5, 81.2, 78.4, 77.9, 76.5, 78.1, 81.8, 79.2, 77.8, 76.9, 
         76.5, 77.4, 78.2, 77.9, 78.7, 77.8, 77.4, 78.1, 77.9, 77.3, 78.3)
nba <- 2*nba
acb <- 2*acb
```


We see now summary statistics of this data:

```{r}
summary(nba)
```

The mean of points in the last twenty years in the NBA is $206.8$ points. The minimum is $186.8$ (in the 2003-04 season) and the maximum is $229.4$, which is the last season.

```{r}
summary(acb)
```

The mean of points in the last twenty years in the ACB is $156.8$. The minimum is $153.0$ (in the 2013-14 season) and the maximum is $167.0$ points (in the current season).

The chosen parameters after solving the system of equations in the NBA case:

```{r}
a_1 <- 17322244/45369
b_1 <- 83240/45369
```

Let us take a look in the probability between the desired values:

```{r}
pgamma(229.4, shape = a_1, rate = b_1)-pgamma(186.8, shape = a_1, rate = b_1)
```

Chosen parameters for the ACB case:

```{r}
a_2 <- 102400/49
b_2 <- 640/49
```

The probability between the desired values:

```{r}
pgamma(167, shape = a_2, rate = b_2) - pgamma(153, shape = a_2, rate = b_2)
```

## Section b)

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
```


Let us read the data:

```{r}
basket <- read.table("basquet.txt", header = TRUE)
```

The priors:

```{r}
## Prior distributions
delta_p <- 0.01
lambda <- seq(120, 250, delta_p)

# NBA prior:
prior_par_nba <- c(shape = a_1, rate = b_1)
prior_nba <- dgamma(lambda, shape=prior_par_nba[1], rate=prior_par_nba[2])

# ACB prior:
prior_par_acb <- c(shape=a_2, rate = b_2)
prior_acb <- dgamma(lambda, shape = prior_par_acb[1], rate = prior_par_acb[2])

# Plot prior comparison
df <- tibble(lambda, prior_nba, prior_acb)
ggplot(df) +
  geom_line(aes(x = lambda, y = prior_nba, linetype = "c")) +
  geom_line(aes(x = lambda, y = prior_acb, linetype = "e")) +
  ggtitle("Prior distribution") +
  ylab("density") +
  scale_linetype_manual(name = "League", 
    values = 1:2, labels = c("NBA", "ACB"))
```

Now the priors likelihoods and posteriors together:

```{r}
## Likelihood distributions
likelihood_nba <- map_dbl(lambda, ~ prod(dpois(basket$nba, .x)))
likelihood_acb <- map_dbl(lambda, ~ prod(dpois(basket$acb, .x)))

## Function to standardize
std_dist <- function(x, delta) {
  x / (sum(x) * delta)
}

## Posterior distributions
# Grid solution:
# Standardize likelihoods
# Calculate products
# Posteriors: standardize products
df <- df %>% 
  mutate(likelihood_nba = std_dist(likelihood_nba, delta_p), 
    likelihood_acb = std_dist(likelihood_acb, delta_p),
    product_nba = prior_nba * likelihood_nba,
    product_acb = prior_acb * likelihood_acb, 
    posterior_nba = std_dist(product_nba, delta_p),
    posterior_acb = std_dist(product_acb, delta_p))

#Plot:
df_long <- df %>% 
    select(-c(product_nba, product_acb)) %>% 
    gather(distribution, density, -lambda) %>% 
    separate(distribution, into = c("distribution", "league"), sep = "_")
 
ggplot(df_long) +
  geom_line(aes(x = lambda, y = density, linetype = league, col = distribution))
```

In the code above we have used the grid solution, one can use the theoretical solutions using that we have worked with the conjugate distribution, and then the posterior is well-known.

## Section c)

We use the simulation approach to simulate the posterior predictive distributions for each league:

```{r}
M <- 1000000
post_sim_nba <- sample(x=lambda, size = M, replace = TRUE, prob = df$posterior_nba)
post_sim_acb <- sample(x=lambda, size = M, replace = TRUE, prob = df$posterior_acb)
pred_post_sim_nba <- rpois(M, post_sim_nba)
pred_post_sim_acb <- rpois(M, post_sim_acb)
gam <- pred_post_sim_acb - pred_post_sim_nba
mean(gam>0)
```

Now let us plot it 

```{r}
ggplot()+
  geom_density(data = tibble(gam), aes(gam))+
  geom_vline(xintercept = 0, color="red")
```

In addition we can compute:

```{r}
mean(gam)
```

Meaning that in a one random match selected for the ACB and the NBA, on average, the score is -29.16 lower in the ACB than in the NBA.

Theoretical option (see the properties of our conjugate prior on (https://en.wikipedia.org/wiki/Conjugate_prior)):

```{r}
pred_post_sim_nba_theo <- rnbinom(M, size = a_1+sum(basket$nba), 
                             prob = (b_1+length(basket$nba))/(1+b_1+length(basket$nba)))
pred_post_sim_acb_theo <- rnbinom(M, size = a_2+sum(basket$acb), 
                             prob = (b_2+length(basket$acb))/(1+b_2+length(basket$acb)))
g <- pred_post_sim_acb_theo-pred_post_sim_nba_theo
mean(g>0)
```

Now the plot:

```{r}
ggplot()+
  geom_density(data = tibble(g), aes(g))+
  geom_vline(xintercept = 0, color="red")
```


Indeed the results obtained are very similar.

## Section d)

Let us compute the $\nu$ quantity:

```{r}
nu <- pred_post_sim_nba - pred_post_sim_acb - 60
mean(nu>0)
```

And finally the plot:

```{r}
ggplot()+
  geom_density(data = tibble(nu), aes(nu))+
  geom_vline(xintercept = 0, color="red")
```

Theoretical alternative:

```{r}
mean(pred_post_sim_nba_theo-pred_post_sim_acb_theo-60>0)
```

Hence, we have a similar result as expected.
